---
title: "Differential Expression and Pathway Enrichment"
author: "Chomchon Uansiri"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Differential Expression and Pathway Enrichment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 8,
  fig.height = 5,
  dpi        = 96,
  out.width  = "100%"
)
```

## Overview

This article goes from cluster markers to pathways and pseudo-bulk
differential expression.

| Step | Functions |
|------|-----------|
| Cluster markers | `FindTopMarkersAndHeatmap()` |
| Marker tables across conditions | `plot_gene_markers_with_dendro()`, `plot_split_dotplots_by_gene_cluster()` |
| Pathway enrichment | `run_global_gsea()`, `run_global_ora()` |
| Pseudo-bulk DESeq2 visualisation | `generate_volcano_trio()`, `get_top_mixed_genes()`, `generate_and_save_heatmap()` |

```{r libraries}
library(Seurat)
library(OHMmy)
library(dplyr)
library(ggplot2)
source("ohmmy-vignette-helpers.R")

library(msigdbr)   # MSigDB gene sets
library(DESeq2)    # pseudo-bulk differential expression

pbmc <- load_pbmc3k_ohmmy()   # built in "Data Processing, Integration, and Clustering"
out_dir <- file.path(tempdir(), "OHMmy-differential-expression")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

The classic GSEA enrichment plots and the volcano plots additionally use the
suggested packages **enrichplot** and **EnhancedVolcano**. OHMmy calls them
internally, so they only need to be installed.

## 1. Cluster markers with `FindTopMarkersAndHeatmap()`

`FindTopMarkersAndHeatmap()` runs `FindAllMarkers()` on the active identities,
or `FindMarkers()` for a single pairwise comparison when `compare` is given. It
uses `logfc.threshold = 0` and `min.pct = 0`, so the complete ranked table
needed for GSEA is kept. Markers are then ranked by a combined relevance score,
`|log2FC| x |pct.1 - pct.2| / p_adj`, filtered by the three `marker_*`
thresholds, and the top `top_n` per cluster are drawn in a `DoHeatmap`. Three
CSV files are written: all markers, a GSEA/ORA-ready copy, and the filtered top
markers.

```{r markers, results = "hide"}
Idents(pbmc) <- "cell_type"

mk <- FindTopMarkersAndHeatmap(
  seurat_obj      = pbmc,
  sample_name     = "pbmc3k_celltype",
  top_n           = 6,
  onlyPos         = TRUE,
  output_dir_base = file.path(out_dir, "markers"),
  height          = 11,
  dpi             = 100
)
```

```{r markers-show, fig.width = 9, fig.height = 10}
names(mk)
mk$top_markers %>%
  group_by(cluster) %>%
  slice_head(n = 3) %>%
  dplyr::select(cluster, gene, avg_log2FC, pct.1, pct.2, p_val_adj, score)
mk$heatmap
```

### Pairwise comparison

With `compare = c(ident.1, ident.2)`, the function runs a single
`FindMarkers()` test. It adds the `gene` and `cluster` columns (each gene is
assigned to the identity in which it is higher), so the output has the same
format as the all-cluster version. With `onlyPos = FALSE`, genes enriched on
both sides are kept.

```{r markers-pair, results = "hide"}
mk_mono <- FindTopMarkersAndHeatmap(
  seurat_obj      = subset(pbmc, subset = cell_type %in% c("CD14 Mono", "FCGR3A Mono")),
  sample_name     = "pbmc3k_monocytes",
  compare         = c("FCGR3A Mono", "CD14 Mono"),
  onlyPos         = FALSE,
  top_n           = 10,
  output_dir_base = file.path(out_dir, "markers"),
  height          = 7,
  dpi             = 100
)
```

```{r markers-pair-show, fig.width = 8, fig.height = 6}
table(mk_mono$top_markers$cluster)
mk_mono$heatmap
```

## 2. Comparing marker tables across conditions

When markers are computed separately for each condition, batch or
integration method, the question becomes how the marker profiles differ
between those runs. Two functions take such a combined table. It needs the
columns `gene`, `cluster`, `avg_log2FC`, `diff` (= `pct.1 - pct.2`) and an
identifier column.

Here we compute cluster markers separately for each of the six
severity-by-batch groups of the simulated cohort.

```{r markers-by-group}
pbmc$group <- paste(pbmc$severity, pbmc$batch, sep = "_")

marker_tbl <- bind_rows(lapply(sort(unique(pbmc$group)), function(g) {
  sub <- subset(pbmc, subset = group == g)
  Idents(sub) <- "seurat_clusters"
  FindAllMarkers(sub, only.pos = TRUE, logfc.threshold = 0.25, min.pct = 0.1, verbose = FALSE) %>%
    mutate(diff = pct.1 - pct.2, group = g)
}))

marker_tbl %>% dplyr::count(group)
```

### `plot_gene_markers_with_dendro()`

Averages the marker statistics per gene and group, clusters the groups (Ward.D2)
and draws a dot plot. Dot size shows log2FC and colour shows the difference in
detection rate. The clusters in which each gene was a marker are written above
the dots, and the dendrogram is attached beside the plot. `genes` accepts a
vector or a single regular expression.

```{r markers-dendro, results = "hide"}
gm <- plot_gene_markers_with_dendro(
  df         = marker_tbl,
  genes      = c("CD14", "LYZ", "S100A8", "S100A9", "FCGR3A", "MS4A7",
                 "GNLY", "NKG7", "MS4A1", "CD79A", "IL7R", "CCR7", "PPBP"),
  id_col     = "group",
  plot_title = "Canonical markers across cohort groups",
  output_dir = file.path(out_dir, "marker_dendro")
)
```

```{r markers-dendro-show, fig.width = 11, fig.height = 4.5}
gm$combined
```

### `plot_split_dotplots_by_gene_cluster()`

For gene families with many members, this function plots every
*gene x cluster* combination. It splits the x-axis into `n_splits` dot plots and
saves a global dendrogram of the groups. Here we use a regular expression to
select all HLA class II genes.

```{r split-dot, results = "hide"}
plot_split_dotplots_by_gene_cluster(
  df                = marker_tbl,
  gene_regex        = "^HLA-D",
  id_col            = "group",
  n_splits          = 2,
  plot_title_prefix = "HLA class II markers",
  output_dir        = file.path(out_dir, "split_dot")
)
```

```{r split-dot-show, echo = FALSE}
show_saved(file.path(out_dir, "split_dot"), "^DotPlot_Part1")
show_saved(file.path(out_dir, "split_dot"), "^Dendrogram_ALL", max_px = 1200)
```

## 3. Pathway enrichment

Both enrichment pipelines take a marker table with one block of genes per
`cluster` and a two-column **TERM2GENE** table, here the MSigDB Hallmark
collection. If the table also has a column whose name contains `cell_type`,
results are additionally summarised per cell type. We therefore compute markers
for the nine `seurat_clusters` (two of which are CD4 T) and attach their cell
type.

```{r gsea-input, results = "hide"}
hallmark <- msigdbr(species = "Homo sapiens", collection = "H")[, c("gs_name", "gene_symbol")]

Idents(pbmc) <- "seurat_clusters"
mk_clusters <- FindTopMarkersAndHeatmap(
  seurat_obj      = pbmc,
  sample_name     = "pbmc3k_clusters",
  onlyPos         = FALSE,             # GSEA needs both tails of the ranking
  output_dir_base = file.path(out_dir, "markers"),
  dpi             = 72
)

enrich_df <- mk_clusters$markers %>%
  mutate(cluster   = as.character(cluster),
         cell_type = as.character(pbmc$cell_type[match(cluster, pbmc$seurat_clusters)]))
```

```{r gsea-input-show}
dim(enrich_df)
dplyr::as_tibble(dplyr::distinct(enrich_df, cluster, cell_type))
```

### Gene Set Enrichment Analysis: `run_global_gsea()`

For every cluster, genes are ranked by `avg_log2FC` and tested with
`clusterProfiler::GSEA()`. A first pass collects the NES range of all clusters,
so every per-cluster dot plot shares the same x-axis and can be compared
directly. The function saves per-cluster NES dot plots, a classic enrichment
plot for the top pathway of each cluster, alphabetical and hierarchically
clustered overview dot plots, signed-significance plots (global and per cell
type), and CSV summaries.

```{r gsea, results = "hide"}
gsea_dir <- file.path(out_dir, "gsea")

gsea_res <- run_global_gsea(
  GSEA_df             = enrich_df,
  m_t2g               = hallmark,
  output_dir          = gsea_dir,
  title_prefix        = "Hallmark",
  top_n_per_direction = 8,
  top_n_overall       = 4
)
```

```{r gsea-table}
gsea_res %>%
  group_by(cluster) %>%
  slice_min(p.adjust, n = 2, with_ties = FALSE) %>%
  dplyr::select(cluster, cell_type, Description, NES, p.adjust) %>%
  arrange(as.numeric(cluster))
```

The clustered overview groups clusters with similar pathway activity. The two
CD4 T clusters (0 and 1) are neighbours, while CD14 monocytes (cluster 2) and
platelets (cluster 8) stand apart. Cluster 2 has activated inflammatory and
TNF-alpha signalling, and cluster 8 has activated coagulation. Clusters without
significant pathways, here the small DC cluster 7, are left out.

```{r gsea-clustered, echo = FALSE}
show_saved(gsea_dir, "Hallmark_GSEA_Clustered_Dendro_Dotplot")
```

The signed-significance view puts direction and significance on one axis:

```{r gsea-signed, echo = FALSE}
show_saved(gsea_dir, "Hallmark_GSEA_Signed_Significance_Dotplot")
```

Per-cluster outputs for cluster 2 (CD14 monocytes): the NES dot plot and the
running enrichment score of its top pathway.

```{r gsea-cluster, echo = FALSE, out.width = "49%", fig.show = "hold"}
show_saved(gsea_dir, "^Hallmark_2_GSEA_NES_Dotplot")
show_saved(gsea_dir, "^Hallmark_2_.*_Gseaplot")
```

### Over-Representation Analysis: `run_global_ora()`

ORA splits each cluster's genes into **Activated** (`avg_log2FC > log2fc_cutoff`)
and **Suppressed** (`avg_log2FC < -log2fc_cutoff`) sets at `p_val_adj <
padj_cutoff`, and tests each set with `clusterProfiler::enricher()` against the
universe of all tested genes. With `variable_per_clus = TRUE`, the function
also draws a signed-significance plot for every cell type.

```{r ora, results = "hide"}
ora_dir <- file.path(out_dir, "ora")

ora_res <- run_global_ora(
  ORA_df            = enrich_df,
  m_t2g             = hallmark,
  output_dir        = ora_dir,
  title_prefix      = "Hallmark",
  top_n_global      = 20,
  top_n_per_cluster = 8,
  log2fc_cutoff     = 0.25,
  padj_cutoff       = 0.05,
  global_width      = 11,
  global_height     = 8,
  variable_per_clus = TRUE
)
```

```{r ora-table}
ora_res %>%
  dplyr::count(cluster, cell_type, Direction) %>%
  arrange(as.numeric(cluster))
```

```{r ora-show, echo = FALSE}
show_saved(ora_dir, "Hallmark_Global_Activated_Clustered_Dendro")
show_saved(ora_dir, "Hallmark_ORA_Signed_Significance_Dotplot")
```

Per-cluster ORA outputs (cluster 2, CD14 monocytes):

```{r ora-cluster, echo = FALSE, out.width = "49%", fig.show = "hold"}
show_saved(ora_dir, "^Hallmark_2_Dotplot")
show_saved(ora_dir, "^Hallmark_2_Barplot")
```

## 4. Pseudo-bulk differential expression with DESeq2

Cluster markers compare cells. A pseudo-bulk analysis compares **samples**: the
counts of each donor x population are summed and analysed with DESeq2, which
correctly models donor-to-donor variability. Here we compare FCGR3A+ with CD14+
monocytes within the same donors, using a paired design (`~ donor + subset`).

```{r pseudobulk}
mono <- subset(pbmc, subset = cell_type %in% c("CD14 Mono", "FCGR3A Mono"))
mono$subset <- ifelse(mono$cell_type == "CD14 Mono", "CD14", "FCGR3A")

pb <- AggregateExpression(mono, assays = "RNA", group.by = c("donor_id", "subset"),
                          return.seurat = FALSE)$RNA
coldata <- data.frame(
  row.names = colnames(pb),
  donor     = sub("_.*$", "", colnames(pb)),
  subset    = factor(sub("^.*_", "", colnames(pb)), levels = c("CD14", "FCGR3A"))
)
coldata$severity <- pbmc$severity[match(coldata$donor, pbmc$donor_id)]

dds <- DESeqDataSetFromMatrix(round(pb), colData = coldata, design = ~ donor + subset)
dds <- dds[rowSums(counts(dds) >= 10) >= 3, ]
dds <- DESeq(dds, quiet = TRUE)

res_df <- as.data.frame(results(dds, contrast = c("subset", "FCGR3A", "CD14")))
head(res_df[order(res_df$padj), c("baseMean", "log2FoldChange", "padj")])
```

### Volcano trio: `generate_volcano_trio()`

The function expects a data frame with gene symbols as row names and the
DESeq2 columns `log2FoldChange` and `padj`. Seurat results can be used after
renaming `avg_log2FC` and `p_val_adj`. It returns three volcano plots (made with
EnhancedVolcano) that label, in turn, the top 10 up- and down-regulated genes,
the significant genes from your own `target_genes` list, and both together.
Genes are labelled only if they pass `padj < 0.05` and `|log2FC| >= 1`.

```{r volcano, fig.width = 20, fig.height = 7.5}
volcano <- generate_volcano_trio(
  res          = res_df,
  main_title   = "FCGR3A+ vs CD14+ monocytes (pseudo-bulk, paired by donor)",
  target_genes = c("FCGR3A", "MS4A7", "CDKN1C", "LST1", "CSF1R",
                   "CD14", "LYZ", "S100A8", "VCAN", "CCR2")
)
volcano
```

### Heatmap of the top genes: `get_top_mixed_genes()` and `generate_and_save_heatmap()`

`get_top_mixed_genes()` returns the union of the `n_padj` most significant genes
and the `n_lfc` genes with the largest effect, chosen among the strictly
significant ones. This way neither highly significant small changes nor large
effects in lowly expressed genes are missed.

```{r top-mixed}
top_genes <- get_top_mixed_genes(res_df, n_padj = 25, n_lfc = 25)
length(top_genes)
head(top_genes, 15)
```

`generate_and_save_heatmap()` uses these genes to draw two row-scaled heatmaps
of the variance-stabilised counts. The left heatmap keeps your sample order
(`ordered_samps`) and the right one clusters the samples freely.

```{r heatmap, results = "hide"}
vsd <- vst(dds, blind = FALSE)

anno_col <- as.data.frame(colData(vsd)[, c("subset", "severity")])
ordered_samples <- rownames(anno_col)[order(anno_col$subset, anno_col$severity)]

heatmap_dir <- file.path(out_dir, "deseq2")

generate_and_save_heatmap(
  res_obj       = res_df,
  comp_name     = "FCGR3A_vs_CD14",
  comp_title    = "FCGR3A+ vs CD14+ monocytes",
  vsd_data      = vsd,
  anno_col      = anno_col,
  ordered_samps = ordered_samples,
  n_padj        = 25,
  n_lfc         = 25,
  out_dir       = heatmap_dir,
  ts            = format(Sys.time(), "%Y%m%d_%H%M%S"),
  clus          = "Monocytes"
)
```

```{r heatmap-show, echo = FALSE}
show_saved(heatmap_dir, "^Heatmap_FCGR3A_vs_CD14")
```

In the unordered (right) heatmap the samples split by monocyte subset, not by
the simulated severity. This is expected, because severity was simulated and
has no transcriptional effect.

## Session information

```{r session-info}
sessionInfo()
```
