---
title: "Ambient RNA Decontamination with SoupX"
author: "Chomchon Uansiri"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Ambient RNA Decontamination with SoupX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 8,
  fig.height = 5,
  dpi        = 96,
  out.width  = "100%"
)
```

## Overview

In droplet-based scRNA-seq, transcripts released by lysed cells form a "soup"
that is captured in every droplet. This ambient RNA makes highly expressed
genes, such as monocyte *LYZ* and *S100A8/9* or haemoglobin, appear in
unrelated cell types. [SoupX](https://github.com/constantAmateur/SoupX)
estimates the soup profile from empty droplets, estimates the contamination
fraction (*rho*) per sample, and subtracts the soup from each cell.

OHMmy wraps SoupX into a multi-sample pipeline that reads Cell Ranger output
directories directly:

| Level | Functions |
|-------|-----------|
| Building blocks | `get_sample_names()`, `load_counts()`, `create_soup_channels()`, `create_seurat_for_clustering()`, `estimate_contamination()`, `create_final_seurat()` |
| Two-stage pipeline | `prepare_soupx_inputs()` (run once) + `run_soupx_post_clustering()` (run per parameter setting) |
| One-step wrapper | `process_soupx_samples()` |
| Hand-off to Seurat | `addSoupXMetaToSeurat()` |

```{r libraries}
library(Seurat)
library(OHMmy)
library(ggplot2)
source("ohmmy-vignette-helpers.R")
```

## 1. Input data

OHMmy expects the standard Cell Ranger layout, with one folder per sample in
each of two parent directories:

```
filtered/<sample>_filtered_feature_bc_matrix/{matrix.mtx.gz, features.tsv.gz, barcodes.tsv.gz}
raw/<sample>_raw_feature_bc_matrix/{matrix.mtx.gz, features.tsv.gz, barcodes.tsv.gz}
```

pbmc3k is only distributed as a filtered matrix. To get a fully reproducible
example, `make_mock_cellranger()` (in `ohmmy-vignette-helpers.R`) splits the
pbmc3k cells into two samples, `PBMC_A` and `PBMC_B`. For each sample it
writes a *raw* matrix that also contains 6,000 **simulated empty droplets**,
whose UMIs are drawn from the pooled expression profile of that sample, which
mimics how ambient RNA behaves.

```{r mock-input}
cr <- make_mock_cellranger(file.path(tempdir(), "cellranger"))
list.files(dirname(cr$filtered_dir), recursive = TRUE)
```

## 2. Step-by-step with the building blocks

### Discover samples and load matrices

`get_sample_names()` derives sample IDs from the folder names. `load_counts()`
reads the filtered and raw matrices of every sample.

```{r load, results = "hide"}
samples <- get_sample_names(cr$filtered_dir)
counts  <- load_counts(samples, filtered_dir = cr$filtered_dir, raw_dir = cr$raw_dir)
```

```{r load-show}
samples
sapply(counts$filtered, ncol)   # cells
sapply(counts$raw, ncol)        # all droplets
```

### Build SoupChannels and a quick clustering

`create_soup_channels()` pairs each raw matrix (`tod`, all droplets) with its
filtered matrix (`toc`, cells). SoupX estimates contamination more reliably
with **over-clustered** cells, so `create_seurat_for_clustering()` runs a fast
standard Seurat pipeline per sample at resolution 1.8.

```{r channels, results = "hide"}
soup_list   <- create_soup_channels(counts$raw, counts$filtered, samples)
seurat_list <- create_seurat_for_clustering(counts$filtered, samples)
```

```{r channels-show}
soup_list$PBMC_A
sapply(seurat_list, function(x) nlevels(x$seurat_clusters))
```

### Estimate and remove contamination

`estimate_contamination()` sets the clusters on each channel, runs
`SoupX::autoEstCont()` and adjusts the counts. The contamination fraction can
be raised on purpose with `mulFac`, given as **percentage points** added to the
estimated *rho*:

* `manual = FALSE` (default): `mulFac` is added for **every** sample;
* `manual = TRUE`: `mulFac` is added **only** for the samples listed in `manual_contam`.

```{r estimate, results = "hide", fig.show = "hide"}
est_auto <- estimate_contamination(
  soup_list, seurat_list, counts$filtered, samples,
  mulFac = 0, methods = "subtraction"
)

# PBMC_B looks noisier: add 5 percentage points to its rho only
est_manual <- estimate_contamination(
  soup_list, seurat_list, counts$filtered, samples,
  mulFac = 5, manual = TRUE, manual_contam = "PBMC_B"
)
```

```{r estimate-show}
get_rho <- function(soup_objs) sapply(soup_objs, function(sc) unique(sc$metaData$rho))
rbind(auto = get_rho(est_auto$soup_list_all), selective = get_rho(est_manual$soup_list_all))
```

`autoEstCont()` estimates about 6 % contamination. The selective override raises
only `PBMC_B`. The estimated soup profile shows which genes dominate the
ambient pool:

```{r soup-profile}
sp <- est_auto$soup_list_all$PBMC_A$soupProfile
head(sp[order(-sp$est), ], 10)
```

`create_final_seurat()` turns the corrected channels into new Seurat objects
(with `min.cells = 3`, `min.features = 200`, and a `mitoPercent` column):

```{r final, results = "hide"}
clean_list <- create_final_seurat(est_auto$soup_list_all, samples)
```

```{r final-show}
clean_list
```

## 3. The two-stage pipeline

Loading, channel creation and clustering do not depend on the contamination
settings. `prepare_soupx_inputs()` therefore bundles them and runs them
**once**. `run_soupx_post_clustering()` takes that bundle and only performs
estimation, correction and Seurat creation, so parameter sweeps are cheap.

```{r prepare, results = "hide"}
prep <- prepare_soupx_inputs(cr$filtered_dir, cr$raw_dir)
```

```{r prepare-show}
names(prep)
```

Here we sweep `multiFac` and record how many UMIs remain:

```{r sweep, results = "hide", fig.show = "hide"}
sweep <- lapply(c(0, 5, 10, 20), function(mf) {
  res <- run_soupx_post_clustering(prep, multiFac = mf, methods = "subtraction")
  data.frame(
    multiFac = mf,
    sample   = names(res$final_seurat),
    rho      = get_rho(res$soup_objects),
    umis     = sapply(res$final_seurat, function(x) sum(x$nCount_RNA))
  )
})
sweep <- do.call(rbind, sweep)
```

```{r sweep-show, fig.width = 7, fig.height = 4}
raw_umis <- sapply(counts$filtered, sum)
sweep$pct_removed <- 100 * (1 - sweep$umis / raw_umis[sweep$sample])
sweep

ggplot(sweep, aes(rho, pct_removed, colour = sample)) +
  geom_line() + geom_point(size = 2.5) +
  labs(x = "Contamination fraction (rho)", y = "% of UMIs removed", colour = NULL) +
  theme_bw()
```

### Per-sample overrides

Giving `manual_contam` restricts the `multiFac` penalty to those samples. All
other samples keep their auto-estimated *rho*:

```{r override, results = "hide", fig.show = "hide"}
override <- run_soupx_post_clustering(prep, multiFac = 10, manual_contam = "PBMC_B")
```

```{r override-show}
get_rho(override$soup_objects)
```

### One-step wrapper

When no parameter sweep is needed, `process_soupx_samples()` runs both stages
in a single call and returns the same structure as
`run_soupx_post_clustering()`:

```{r one-step, results = "hide", fig.show = "hide"}
one_step <- process_soupx_samples(cr$filtered_dir, cr$raw_dir, multiFac = 0)
```

```{r one-step-show}
names(one_step)
get_rho(one_step$soup_objects)
```

## 4. Transferring the cleaned counts to an existing object

Often an analysed Seurat object already exists, built from the uncorrected
counts. `addSoupXMetaToSeurat()` merges the per-sample SoupX objects, prefixes
the barcodes with the sample name and removes the `-1` suffix. It then
**replaces the `counts` layer** of the original object and updates
`nCount_RNA`, `nFeature_RNA` and `pct_counts_mt`, keeping only cells and genes
present in both objects.

To show this, we build the "original" object with the same barcode convention
(`<sample>_<barcode>`):

```{r original}
original <- CreateSeuratObject(
  do.call(cbind, lapply(samples, function(s) {
    m <- counts$filtered[[s]]
    colnames(m) <- paste0(s, "_", sub("-1$", "", colnames(m)))
    m
  })),
  project = "pbmc3k_uncorrected"
)
original$sample <- sub("_.*$", "", colnames(original))
head(colnames(original), 3)
```

```{r transfer, results = "hide"}
soupx_run <- run_soupx_post_clustering(prep, multiFac = 0)

corrected <- addSoupXMetaToSeurat(
  original_seurat   = original,
  soupx_seurat_list = soupx_run$final_seurat,
  mito_colname      = "mitoPercent",
  save_path         = NULL   # give a .rds path to save the result
)
```

```{r transfer-show}
corrected
```

### Did it work?

The clearest check is ambient monocyte transcripts in **non-monocytes**. We
compare, before and after correction, the fraction of lymphocytes (cells with
low *LYZ* + *CD14* expression in a marker-based gate) that still contain
monocyte-specific UMIs:

```{r evaluate, fig.width = 9, fig.height = 4}
before <- LayerData(original,  layer = "counts")
after  <- LayerData(corrected, layer = "counts")
cells  <- colnames(after)
before <- before[rownames(after), cells]

# gate: cells whose LYZ share is in the lower 70 % are treated as non-monocytes
lyz_share  <- before["LYZ", ] / Matrix::colSums(before)
non_mono   <- cells[lyz_share < quantile(lyz_share, 0.70)]

ambient_genes <- c("LYZ", "S100A8", "S100A9", "FTL")
detect <- function(m) 100 * Matrix::rowMeans(m[ambient_genes, non_mono] > 0)
data.frame(gene = ambient_genes,
           pct_nonmono_before = round(detect(before), 1),
           pct_nonmono_after  = round(detect(after), 1))

plot_df <- data.frame(
  nCount = c(original$nCount_RNA[cells], corrected$nCount_RNA),
  state  = rep(c("before SoupX", "after SoupX"), each = length(cells))
)
ggplot(plot_df, aes(nCount, fill = state)) +
  geom_density(alpha = 0.5) + scale_x_log10() +
  labs(x = "UMIs per cell (log10)", fill = NULL) + theme_bw()
```

SoupX removes about 6 % of the UMIs, spread across all cells. Its effect on
the ambient transcripts is much larger: *LYZ*, *S100A8* and *S100A9*, which
were detected in a large share of non-monocytes before correction, almost
disappear from them afterwards. Ubiquitously expressed genes such as *FTL* are
barely affected, which is the intended behaviour. `corrected` can now go into
`ProcessSeuratLOG()` or `ProcessSeuratSCT()`. See the article *Data Processing,
Integration, and Clustering*.

## Session information

```{r session-info}
sessionInfo()
```
