---
title: "Comprehensive single-cell workflow with OHMmy"
author: "Chomchon Uansiri"
package: "OHMmy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comprehensive single-cell workflow with OHMmy}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  eval = FALSE # Set to FALSE for fast package building
)
```

## Introduction

`OHMmy` provides a complete, natively integrated toolkit for Seurat v5.
It is designed to handle every step of single-cell RNA-seq analysis:
from raw data ingestion and ambient RNA removal, to advanced
hierarchical visualizations and GSEA enrichment.

This vignette covers the entire lifecycle of an analysis using all
exported functions in the package.

## 1. Data I/O and Quality Control

Start by loading your raw data and performing initial QC checks across
your samples.

```{r}

library(Seurat)
library(OHMmy)

# 1. Load data and extract metadata
counts_list <- load_counts(data_dir = "path/to/data")
sample_names <- get_sample_names(counts_list)

# 2. Visualize initial Quality Control metrics
plot_violin_qc_single(counts_list[[1]], feature = "nFeature_RNA")
```

## 2. The SoupX Decontamination Pipeline

`OHMmy` excels at automating ambient RNA removal. You can run the
pipeline step-by-step, or use the wrapper functions to completely
automate the transition from raw counts to a clean Seurat object.

```{r}

# Step-by-step SoupX preparation
soupx_inputs <- prepare_soupx_inputs(raw_counts = counts_list$raw, filtered_counts = counts_list$filtered)
soup_channels <- create_soup_channels(soupx_inputs$raw, soupx_inputs$filtered)

# Pre-clustering for contamination estimation
seurat_pre <- create_seurat_for_clustering(soupx_inputs$filtered)
soup_channels <- estimate_contamination(soup_channels, clusters = seurat_pre$seurat_clusters)

# Or, use the overarching wrapper to process and build the final clean object
clean_matrices <- process_soupx_samples(soup_channels)
seurat_obj <- create_final_seurat(clean_matrices)

# Post-clustering ambient RNA refinement and metadata tracking
seurat_obj <- run_soupx_post_clustering(seurat_obj)
seurat_obj <- addSoupXMetaToSeurat(seurat_obj, metadata_df)
```

## 3. Core Processing & Clustering

Once your object is clean, choose your preferred normalization method
(Log or SCTransform) and compute dimensionality reductions.

```{r}

# Choose your processing pipeline:
# seurat_obj <- ProcessSeuratLOG(seurat_obj) # Standard log-normalization
seurat_obj <- ProcessSeuratSCT(seurat_obj)   # SCTransform (Recommended)

# Run PCA, UMAP, and clustering
seurat_obj <- ClusterAndUMAP(seurat_obj, resolution = 0.5)

# Clean up any stale or unnecessary reductions to save memory
seurat_obj <- CleanSeuratReductions(seurat_obj)

# Create a custom cohesive color palette for your new clusters
cluster_colors <- make_cluster_palette(seurat_obj)
```

## 4. Diagnostics & Technical Noise Evaluation

Before diving into biology, evaluate if technical artifacts (like cell
cycle or sequencing depth) are driving your clusters or principal
components.

```{r}

# Check PCA loadings and metadata correlations
plot_vizdimloadings(seurat_obj, dims = 1:2)
plot_pc_metadata_correlation(seurat_obj, metadata_cols = c("nCount_RNA", "percent.mt"))

# Assess technical contribution per cluster
plot_technical_contribution(seurat_obj, feature = "percent.mt")
plot_stacked_technical_contribution(seurat_obj, features = c("percent.mt", "percent.ribo"))
```

## 5. Visualizing Clusters & Metadata

Use `OHMmy`'s specialized demographic tools to see how your cells are
distributed across conditions.

```{r}

# Visualize UMAP split by experimental factors
PlotDimByFactors(seurat_obj, group.by = "Condition")
plot_combined(seurat_obj, reduction = "umap", group.by = "Condition")

# Statistical evaluation of cluster distributions and metadata
plot_cluster_distributions(seurat_obj, group.by = "SampleID")
plot_cell_abundance(seurat_obj, group.by = "Condition", test = "wilcox")
plot_metadata_stats(seurat_obj, feature = "score", group.by = "Cluster")
```

## 6. High-Resolution Gene & Marker Visualization

This is where `OHMmy` truly shines. Explore gene expression using
hierarchical dotplots, Nebulosa density blends, and complex heatmaps.

### Density & Co-Expression Blends

```{r}

# Visualize gene pair correlations and overlap
plot_gene_pair_correlations(seurat_obj, feature1 = "CD8A", feature2 = "XCL1")
plot_blend_nebulosa(seurat_obj, features = c("CD8A", "XCL1"))

# Internal V5 blend wrapper for strict layer management
# .blend_feature_plot_v5(seurat_obj, features = c("CD4", "IL2RA"))
```

### Advanced Dotplots (with Dendrograms)

```{r}

# Generate standard dotplots with hierarchical clustering trees
plot_dot_dendro(seurat_obj, features = top_markers)

# Split dotplots by condition or metadata
plot_dot_dendro_split(seurat_obj, features = top_markers, split.by = "Condition")
plot_dot_dendro_multi(seurat_obj, feature_list = list(Tcells = c("CD3D"), Bcells = c("CD79A")))
plot_split_dotplots_by_gene_cluster(seurat_obj, features = top_markers)
plot_gene_markers_with_dendro(seurat_obj, features = top_markers)
```

### Heatmaps & Gene Binning

```{r}

# Extract binned expression and plot top mixed genes
binned_expr <- extract_binned_expression(seurat_obj, bins = 10)
top_mixed <- get_top_mixed_genes(seurat_obj)

# Comprehensive heatmaps
FindTopMarkersAndHeatmap(seurat_obj, group.by = "seurat_clusters")
generate_and_save_heatmap(seurat_obj, features = top_markers, filename = "heatmap.png")
generate_dimheatmaps(seurat_obj, dims = 1:5)
```

## 7. Differential Expression & Functional Enrichment

Extract biological meaning by finding differential genes and running
Over-Representation Analysis (ORA) or Gene Set Enrichment Analysis
(GSEA).

```{r}

# Generate a trio of volcano plots comparing multiple conditions
generate_volcano_trio(seurat_obj, ident.1 = "Infected", ident.2 = "Control")

# Run global enrichment profiling on your markers
# ora_results <- run_global_ora(seurat_obj, database = "GO")
# gsea_results <- run_global_gsea(seurat_obj, database = "KEGG")
```

## Conclusion

The `OHMmy` package minimizes boilerplate code and maximizes analytical
rigor. By combining strict Seurat v5 layer management with advanced
statistical visualizations, it provides a stable foundation for complex
scRNA-seq discoveries.
