---
title: "Data Processing, Integration, and Clustering"
author: "Chomchon Uansiri"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Data Processing, Integration, and Clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 8,
  fig.height = 5,
  dpi        = 96,
  out.width  = "100%"
)
```

## Overview

This article takes the public **10x Genomics 3k PBMC** dataset from a raw count
matrix to an integrated, clustered and annotated Seurat v5 object using the
OHMmy processing wrappers:

| Step | OHMmy function | What it automates |
|------|----------------|-------------------|
| QC inspection | `plot_violin_qc_single()` | Jittered violin plots of QC metrics, saved to disk |
| Normalisation + integration | `ProcessSeuratLOG()` / `ProcessSeuratSCT()` | Layer splitting, HVG selection, TCR/BCR exclusion, regression, PCA, `IntegrateLayers()` with batch-size-aware `k` parameters, elbow plot |
| Clustering + UMAP | `ClusterAndUMAP()` | Neighbour graph, resolution sweep, `clustree` stability plot, final clustering and UMAP |
| Housekeeping | `CleanSeuratReductions()` | Renames reductions such as `pca.log` to Seurat-safe camelCase keys |

The object produced here is the one used by all other OHMmy articles.

## 1. Load the data and compute QC metrics

```{r libraries}
library(Seurat)
library(OHMmy)
library(patchwork)

# Shared helpers: data download/caching, the simulated cohort and figure display
source("ohmmy-vignette-helpers.R")

# SCTransform + anchor-based integration ship large objects to future workers
options(future.globals.maxSize = 4 * 1024^3)

out_dir <- file.path(tempdir(), "OHMmy-processing-clustering")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

`pbmc3k_counts()` reads the filtered feature-barcode matrix, downloading it once
from 10x Genomics if it is not already available locally.

```{r load-data}
pbmc <- CreateSeuratObject(pbmc3k_counts(), project = "pbmc3k", min.cells = 3, min.features = 200)

pbmc[["pct_counts_mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
pbmc[["percent.ribo"]]  <- PercentageFeatureSet(pbmc, pattern = "^RP[SL]")
pbmc
```

### Inspect QC distributions

`plot_violin_qc_single()` overlays faded single-cell points on Seurat violins so
that outliers stay visible. Any metadata column can be used for grouping; before
clustering we simply group by `orig.ident`.

```{r qc-violin, results = "hide"}
plot_violin_qc_single(
  seurat_obj       = pbmc,
  sample_name      = "pbmc3k_preQC",
  qc_meta_features = c("nCount_RNA", "nFeature_RNA", "pct_counts_mt", "percent.ribo"),
  res_col          = "orig.ident",
  output_dir       = out_dir,
  dpi              = 100
)
```

```{r qc-violin-show, echo = FALSE}
show_saved(out_dir, "pbmc3k_preQC_ViolinQC")
```

Based on these distributions we apply the standard pbmc3k thresholds.

```{r qc-filter}
pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & pct_counts_mt < 5)
ncol(pbmc)
```

### A simulated cohort

pbmc3k comes from one donor. So that the batch-integration, abundance and
pseudo-bulk functions have something to work with, `add_mock_cohort()` spreads
the cells across **12 simulated donors** (Healthy / Mild / Severe, 4 each) that
were sequenced in **two batches**. Monocyte-like cells are deliberately
over-assigned to Mild and Severe donors, which gives the statistics articles a
known effect to recover. *All cohort columns are synthetic.*

```{r mock-cohort}
pbmc <- add_mock_cohort(pbmc)

table(pbmc$severity, pbmc$batch)
head(pbmc@meta.data[, c("donor_id", "severity", "batch", "age", "sex", "infection_status")])
```

## 2. Normalisation and batch integration

`ProcessSeuratLOG()` runs the complete Seurat v5 LogNormalize workflow. It:

1. splits the RNA assay into one layer per batch (`batch_col`);
2. log-normalises the data and selects 2,000 variable features;
3. removes TCR/BCR V(D)J genes (`tcr_bcr_patterns`) from the variable features,
   so clustering reflects cell phenotype rather than clonotype;
4. scales the data while regressing `vars_to_regress`;
5. optionally boosts the variance of lineage markers (`boost_genes`,
   `boost_multiplier`), which helps separate look-alike populations such as CD8 T
   and NK cells;
6. runs PCA, saves an elbow plot and suggests a number of PCs;
7. integrates the layers with Harmony, RPCA, CCA or FastMNN, shrinking `k.weight`,
   `k.anchor`, `k.filter` and `k.score` automatically when a batch is small;
8. re-joins the layers.

```{r process-log, results = "hide"}
pbmc <- ProcessSeuratLOG(
  pbmc,
  batch_col             = "batch",
  vars_to_regress       = "pct_counts_mt",
  tcr_bcr_patterns      = "^TR[ABDG]V|^IG[HKL]V",
  reduction_name        = "pca.log",
  integration_method    = "HarmonyIntegration",
  integration_reduction = "harmony.log",
  dims                  = 1:30,
  elbow_plot_dir        = out_dir,
  sample_name           = "pbmc3k",
  verbose               = FALSE
)
```

```{r elbow-show, echo = FALSE, out.width = "70%"}
show_saved(out_dir, "^elbow_plot_pbmc3k")
```

```{r process-log-result}
pbmc
Reductions(pbmc)
```

> **Interactive use.** With `interactive_mode = TRUE` in an interactive R
> session, the function pauses after the PCA step so you can choose the number of
> PCs and the integration `k` parameters from the elbow plot.

### Cell-cycle scores

Cell-cycle scores are useful covariates for the PC diagnostics in the
*Diagnostic and Variable Evaluation* article.

```{r cell-cycle}
pbmc <- CellCycleScoring(
  pbmc,
  s.features   = cc.genes.updated.2019$s.genes,
  g2m.features = cc.genes.updated.2019$g2m.genes
)
table(pbmc$Phase)
```

### Alternative: SCTransform workflow

`ProcessSeuratSCT()` has the same interface but normalises with `SCTransform`.
Here it is combined with reciprocal-PCA integration to show a second
integration back-end. It runs on a copy, so the LogNormalize object is
unchanged.

```{r process-sct, results = "hide"}
pbmc_sct <- ProcessSeuratSCT(
  pbmc,
  batch_col             = "batch",
  vars_to_regress       = "pct_counts_mt",
  reduction_name        = "pca.sct",
  integration_method    = "RPCAIntegration",
  integration_reduction = "rpca.sct",
  dims                  = 1:30,
  verbose               = FALSE
)
pbmc_sct <- RunUMAP(pbmc_sct, reduction = "rpca.sct", dims = 1:20,
                    reduction.name = "umap.rpca.sct", verbose = FALSE)
```

```{r sct-compare, fig.width = 11, fig.height = 4.5}
pbmc_tmp <- RunUMAP(pbmc, reduction = "pca.log", dims = 1:20,
                    reduction.name = "umap.unintegrated", verbose = FALSE)

DimPlot(pbmc_tmp, reduction = "umap.unintegrated", group.by = "batch") +
  ggplot2::ggtitle("LogNormalize, no integration") +
  DimPlot(pbmc_sct, reduction = "umap.rpca.sct", group.by = "batch") +
  ggplot2::ggtitle("SCTransform + RPCA") +
  plot_layout(guides = "collect")
```

Because the batches in this example are simulated, both embeddings mix the
batches. In real data the integrated embedding should mix batches while keeping
cell types separate.

```{r cleanup-sct, include = FALSE}
rm(pbmc_sct, pbmc_tmp); invisible(gc())
```

## 3. Clustering across resolutions

`ClusterAndUMAP()` builds the shared-nearest-neighbour graph, clusters across a
range of resolutions, draws a `clustree` plot to show how clusters split as the
resolution increases, then runs the final clustering and UMAP. Graphs and
resolutions that already exist are reused unless `force_neighbors` or
`force_clustering` is `TRUE`.

```{r cluster-umap, results = "hide"}
clus <- ClusterAndUMAP(
  seurat_obj          = pbmc,
  sample_name         = "pbmc3k",
  reduction           = "harmony.log",
  dims                = 1:20,
  k.param             = 20,
  umap_name           = "umap.har",
  graph_name          = "har_snn",
  cluster_prefix      = "har_snn_res.",
  cluster_resolutions = seq(0.2, 1.2, by = 0.2),
  final_resolution    = 0.6,
  plot_dir            = out_dir,
  dpi                 = 100
)
pbmc <- clus$seurat
```

The function returns the updated object, the `clustree` ggplot and the path of
the saved figure:

```{r clustree, fig.width = 8, fig.height = 8}
names(clus)
clus$clustree
```

Between resolutions 0.6 and 1.0 almost every cluster has a single parent,
so the partition is stable. Further splits appear only at 1.2. We use 0.6,
where the major PBMC lineages are already resolved, as the working resolution.

```{r final-clusters}
pbmc$seurat_clusters <- pbmc$har_snn_res.0.6
table(pbmc$seurat_clusters)
```

## 4. Annotation

For a self-contained tutorial, clusters are labelled from canonical PBMC
markers: each cluster receives the cell type whose module score
(`Seurat::AddModuleScore()`) is highest. The helper `annotate_pbmc()` is defined
in `ohmmy-vignette-helpers.R`. In your own data, replace this step with your
preferred annotation strategy, such as reference mapping or manual curation.

```{r annotate}
str(pbmc_marker_sets)
pbmc <- annotate_pbmc(pbmc, cluster_col = "seurat_clusters")
table(pbmc$seurat_clusters, pbmc$cell_type)
```

```{r umap-annotated, fig.width = 11, fig.height = 4.5}
DimPlot(pbmc, reduction = "umap.har", group.by = "seurat_clusters", label = TRUE) + NoLegend() +
  DimPlot(pbmc, reduction = "umap.har", group.by = "cell_type", label = TRUE, repel = TRUE) + NoLegend()
```

## 5. Standardising reduction names

Reductions named with dots or underscores (such as `pca.log`) are valid in Seurat
v5, but some tools expect keys that match the reduction name exactly.
`CleanSeuratReductions()` rebuilds every reduction under a camelCase name with
a matching key. It is shown on a copy, because the remaining articles use the
original names.

```{r clean-reductions}
pbmc_clean <- CleanSeuratReductions(pbmc)
data.frame(
  before = Reductions(pbmc),
  after  = Reductions(pbmc_clean)[match(c("pcaLog", "harmonyLog", "umapHar"), Reductions(pbmc_clean))]
)
Key(pbmc_clean[["umapHar"]])
```

## 6. Saving the object

Every other OHMmy article reuses this object. `load_pbmc3k_ohmmy()` (in
`ohmmy-vignette-helpers.R`) runs exactly the steps above and caches the result,
so later articles start immediately. To save your own object, use `saveRDS()` or
the `save_path` argument of `ClusterAndUMAP()`.

```{r save, eval = FALSE}
saveRDS(pbmc, "pbmc3k_processed.rds")
```

## Session information

```{r session-info}
sessionInfo()
```
