---
title: "Advanced Visualization and Marker Analysis"
author: "Chomchon Uansiri"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Advanced Visualization and Marker Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 8,
  fig.height = 5,
  dpi        = 96,
  out.width  = "100%"
)
```

## Overview

Once cells are clustered, the next step is to interpret them. OHMmy wraps the
most common figures into functions that save a publication-ready file **and**
return the ggplot/patchwork object, so you can keep customising in R.

| Theme | Functions |
|-------|-----------|
| Colours | `make_cluster_palette()` |
| Embedding overlays | `PlotDimByFactors()`, `plot_combined()` |
| Distributions | `plot_violin_qc_single()`, `extract_binned_expression()` |
| Clustered dot plots | `plot_dot_dendro()`, `plot_dot_dendro_split()`, `plot_dot_dendro_multi()` |
| Co-expression | `plot_blend_nebulosa()`, `.blend_feature_plot_v5()` |
| Lineage resolution | `score_consensus()` |

```{r libraries}
library(Seurat)
library(OHMmy)
library(patchwork)
library(ggplot2)
source("ohmmy-vignette-helpers.R")

pbmc <- load_pbmc3k_ohmmy()   # built in "Data Processing, Integration, and Clustering"
out_dir <- file.path(tempdir(), "OHMmy-advanced-visualization")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

table(pbmc$cell_type)
```

## 1. Colour palettes for many clusters

`make_cluster_palette()` returns `n` maximally distinct colours from curated
qualitative palettes (`"kelly"`, `"alphabet"`, `"polychrome"`) or from a blend of
RColorBrewer sets (`"auto"`). When `n` is larger than the palette, it
interpolates.

```{r palettes, fig.width = 9, fig.height = 2.6}
pal_df <- do.call(rbind, lapply(c("auto", "kelly", "alphabet", "polychrome"), function(p) {
  data.frame(palette = p, i = 1:16, col = make_cluster_palette(16, palette = p))
}))
ggplot(pal_df, aes(i, palette, fill = col)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  scale_fill_identity() +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), panel.grid = element_blank())
```

## 2. Embedding overlays

### One UMAP per metadata column

`PlotDimByFactors()` draws one `DimPlot` per metadata column, saves each to
disk and returns the plots as a named list.

```{r dim-factors, results = "hide"}
dim_plots <- PlotDimByFactors(
  seurat_obj  = pbmc,
  factors     = c("cell_type", "seurat_clusters", "batch", "Phase"),
  sample_name = "pbmc3k",
  reduction   = "umap.har",
  output_dir  = file.path(out_dir, "dimplots"),
  dpi         = 100,
  verbose     = FALSE
)
```

```{r dim-factors-show, fig.width = 12, fig.height = 10}
names(dim_plots)
dim_plots$cell_type <- dim_plots$cell_type +
  scale_colour_manual(values = make_cluster_palette(nlevels(pbmc$cell_type), "kelly"))
wrap_plots(dim_plots, ncol = 2) & NoLegend()
```

### Feature plots paired with density estimates

Sparse genes are hard to read on a standard `FeaturePlot`. `plot_combined()`
places each feature next to its kernel-density estimate (Nebulosa), using a
Seurat v5-safe implementation. Features can be **genes or numeric metadata
columns**, and each list element becomes one saved figure.

```{r combined, results = "hide"}
feature_sets <- list(
  Lymphoid = c("CD3E", "MS4A1", "GNLY"),
  Myeloid  = c("CD14", "FCGR3A", "FCER1A"),
  QC       = c("nCount_RNA", "pct_counts_mt", "percent.ribo")
)

combined <- plot_combined(
  seurat_obj  = pbmc,
  cluster_col = "cell_type",
  genes_list  = feature_sets,
  sample_name = "pbmc3k",
  reduction   = "umap.har",
  output_dir  = file.path(out_dir, "combined"),
  dpi         = 72,
  verbose     = FALSE
)
```

```{r combined-show, fig.width = 18, fig.height = 3.6}
names(combined$combined_plots)
combined$combined_plots$Lymphoid
combined$combined_plots$QC
```

## 3. Distributions across clusters

### Violin plots for QC metrics and genes

`plot_violin_qc_single()` saves one figure for QC metadata and one for genes,
grouped by any identity column.

```{r violins, results = "hide"}
plot_violin_qc_single(
  seurat_obj       = pbmc,
  sample_name      = "pbmc3k",
  qc_meta_features = c("nCount_RNA", "nFeature_RNA", "pct_counts_mt"),
  gene_features    = c("CD3E", "CD8A", "GNLY", "MS4A1", "CD14", "FCGR3A"),
  res_col          = "cell_type",
  output_dir       = file.path(out_dir, "violins"),
  dpi              = 90
)
```

```{r violins-show, echo = FALSE}
show_saved(file.path(out_dir, "violins"), "ViolinQC")
show_saved(file.path(out_dir, "violins"), "ViolinGene")
```

### Discretising expression into Low / Int / High

`extract_binned_expression()` returns the average expression and percent
expressed per group. For each gene it then bins the groups into terciles
(`Low`, `Int`, `High`). This is useful for writing marker tables or for defining
"marker-high" populations. The terciles are relative to the groups being
compared: with eight cell types, for example, the three highest-expressing
groups of each gene are labelled `High`.

```{r binned}
binned <- extract_binned_expression(
  seurat_obj = pbmc,
  gene_list  = c("CD3E", "CD8A", "GNLY", "MS4A1", "CD14", "FCGR3A", "FCER1A", "PPBP"),
  group_col  = "cell_type"
)
head(binned)
```

```{r binned-plot, fig.width = 8, fig.height = 4}
ggplot(binned, aes(Gene, Cluster, fill = Expression_Level)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = round(PctExpress)), size = 3) +
  scale_fill_manual(values = c(Low = "#f0f0f0", Int = "#9ecae1", High = "#08519c")) +
  labs(x = NULL, y = NULL, fill = "Tercile", caption = "Numbers: % of cells expressing") +
  theme_minimal()
```

## 4. Hierarchically clustered dot plots

The dot-plot family takes a two-column data frame, `feature_df`, that maps genes
(`feature`) to panels (`group`). Groups are matched by **prefix**: with
`prefix = "PBMC"`, every group starting with `"PBMC_"` is plotted. This lets one
table hold several marker panels, for example `PBMC_*` and `Tcell_*`.

```{r feature-df}
feature_df <- data.frame(
  group   = paste0("PBMC_", rep(names(pbmc_marker_sets), lengths(pbmc_marker_sets))),
  feature = unlist(pbmc_marker_sets, use.names = FALSE)
)
head(feature_df, 8)
```

### `plot_dot_dendro()`: genes and clusters clustered together

Genes expressed in fewer than `pct_threshold` % of cells in every group are
dropped. Rows and columns are then ordered by hierarchical clustering, and both
dendrograms are drawn. The function also saves standalone dendrograms and a
text file of the clustered gene order.

```{r dot-dendro, results = "hide"}
p_dot <- plot_dot_dendro(
  seurat_obj    = pbmc,
  meta_col      = "cell_type",
  feature_df    = feature_df,
  prefix        = "PBMC",
  pct_threshold = 10,
  output_dir    = file.path(out_dir, "dot_dendro")
)
```

```{r dot-dendro-show, fig.width = 12, fig.height = 7.5}
p_dot
```

### `plot_dot_dendro_split()`: large panels in chunks

With long marker lists, `plot_dot_dendro_split()` keeps the **global**
clustering but splits the gene axis into chunks of `max_genes_per_plot`. The
same cluster dendrogram is attached to every chunk.

```{r dot-split, results = "hide"}
extended_markers <- list(
  T_cell   = c("CD3D", "CD3E", "CD2", "IL7R", "CCR7", "SELL", "LEF1", "TCF7", "CD8A", "CD8B", "GZMK"),
  NK       = c("GNLY", "NKG7", "KLRD1", "KLRF1", "PRF1", "GZMB", "GZMH", "FCGR3A"),
  B        = c("MS4A1", "CD79A", "CD79B", "BANK1", "TCL1A", "HLA-DQA1"),
  Myeloid  = c("CD14", "LYZ", "S100A8", "S100A9", "VCAN", "MS4A7", "LST1", "CDKN1C", "FCER1A", "CST3"),
  Platelet = c("PPBP", "PF4", "GNG11")
)
feature_df_ext <- data.frame(
  group   = paste0("PBMCext_", rep(names(extended_markers), lengths(extended_markers))),
  feature = unlist(extended_markers, use.names = FALSE)
)

split_plots <- plot_dot_dendro_split(
  seurat_obj         = pbmc,
  meta_col           = "cell_type",
  feature_df         = feature_df_ext,
  prefix             = "PBMCext",
  pct_threshold      = 10,
  max_genes_per_plot = 20,
  output_dir         = file.path(out_dir, "dot_split")
)
```

```{r dot-split-show, fig.width = 10, fig.height = 5}
length(split_plots)
split_plots[[1]]
split_plots[[2]]
```

### `plot_dot_dendro_multi()`: several groupings in one clustering

`plot_dot_dendro_multi()` stacks the dot-plot summaries of **several metadata
columns** into one matrix before clustering. Here, plotting the marker-based
`cell_type` labels next to the unsupervised `seurat_clusters` shows directly
which clusters each label absorbed. Rows are named `<group>_<column>`.

```{r dot-multi, results = "hide"}
multi_df <- plot_dot_dendro_multi(
  seurat_obj    = pbmc,
  meta_cols     = c("cell_type", "seurat_clusters"),
  feature_df    = feature_df,
  prefix        = "PBMC",
  pct_threshold = 10,
  output_dir    = file.path(out_dir, "dot_multi"),
  save_options  = c("combined", "gene_list")
)
```

```{r dot-multi-show, echo = FALSE}
show_saved(file.path(out_dir, "dot_multi"), "^DotPlot_RowColDendro_PBMC")
```

```{r dot-multi-data}
head(multi_df[, c("cluster_var", "feature", "avg.exp.scaled", "pct.exp")])
```

## 5. Co-expression: blend and density plots

`plot_blend_nebulosa()` produces one row per gene pair with the two single-gene
panels, their colour blend, a 2-D colour key, and a Nebulosa density estimate
for each gene. Cells below `blend_threshold` (after scaling to 0 to 1) are shown
as background, which keeps sparse genes readable. All pairs are stacked into one
saved figure.

```{r blend, results = "hide"}
blend <- plot_blend_nebulosa(
  seurat_obj      = pbmc,
  cluster_col     = "cell_type",
  gene_pairs      = list(c("CD3E", "GNLY"), c("CD14", "FCGR3A")),
  sample_name     = "pbmc3k",
  reduction       = "umap.har",
  blend_threshold = 0.1,
  width           = 24,
  dpi             = 72,
  output_dir      = file.path(out_dir, "blend"),
  verbose         = FALSE
)
```

```{r blend-show, fig.width = 24, fig.height = 7.5}
blend$dimensions
blend$plot
```

The *CD3E*/*GNLY* row separates T cells (green), NK cells (red) and the
cytotoxic CD8 T cells that express both (yellow). The *CD14*/*FCGR3A* row shows
the two monocyte subsets and the intermediate cells between them.

The four-panel blend is also available on its own as `.blend_feature_plot_v5()`,
a Seurat v5-compatible replacement for `FeaturePlot(blend = TRUE)`.

```{r blend-single, fig.width = 14, fig.height = 3.8}
.blend_feature_plot_v5(
  pbmc, gene1 = "CD8A", gene2 = "GZMK", reduction = "umap.har",
  cols = c("lightgrey", "#1b9e77", "#d95f02"), blend_threshold = 0.15, pt_size = 0.6
)
```

## 6. Resolving CD8 T cells from NK cells with `score_consensus()`

Cytotoxic CD8 T cells and NK cells share much of their effector programme and
are often mixed in clustering. `score_consensus()` compares one or more **CD8
signature scores** with an **NK signature score** for each cell. It labels a
cell CD8 when `z(CD8) - z(NK)` exceeds a cutoff, which can be fixed
(`"fixed"`), the density valley (`"trough"`), or the decision boundary of a
two-component Gaussian mixture (`"gmm"`). Two CD8 signatures act as `primary`
and `confirm` scores, and disagreements are flagged as ambiguous. Bootstrapping
reports how stable the cutoff is.

Any per-cell score can be used, such as `AddModuleScore()` or UCell. Here we
score the CD8 T and NK cells with `AddModuleScore()`.

```{r consensus-scores}
cyto <- subset(pbmc, subset = cell_type %in% c("CD8 T", "NK"))

score_sets <- list(
  CD8_core = c("CD8A", "CD8B"),
  CD8_TCR  = c("CD3D", "CD3E", "CD3G", "CD8A", "CD8B", "LCK"),
  NK_score = c("GNLY", "KLRD1", "KLRF1", "FCGR3A", "NCAM1", "TYROBP", "FCER1G")
)
for (s in names(score_sets)) {
  cyto <- AddModuleScore(cyto, features = list(score_sets[[s]]), name = s, seed = 1)
  names(cyto@meta.data)[names(cyto@meta.data) == paste0(s, "1")] <- s
}
```

With `method = "gmm"`, the cutoff is the decision boundary of a two-component
Gaussian mixture fitted with **mclust**, which only needs to be installed. If
no cutoff can be estimated, `fixed_cut` is used instead, and the `fallback`
column of `cons$cutoffs` reports this.

```{r consensus-run}
cons <- score_consensus(
  obj         = cyto,
  cd8_cols    = c("CD8_TCR", "CD8_core"),
  nk_col      = "NK_score",
  method      = "gmm",
  n_boot      = 100,
  boot_target = "cutoff",
  make_plots  = TRUE,
  plot_n      = 300,
  out_dir     = file.path(out_dir, "cd8_vs_nk")
)

cons$cutoffs
cons$agreement
cons$bootstrap$summary
```

Both variants get a data-driven cutoff (`fallback = FALSE`), and the
bootstrap confidence intervals show how stable it is. The consensus agrees
with the cluster-level annotation for the large majority of cells. Most of the
remaining cells are flagged as *Ambiguous*, because the primary and
confirmatory CD8 signatures disagree on them. These cells are good candidates
for closer inspection:

```{r consensus-table}
table(consensus = cons$consensus, annotation = cyto$cell_type)
```

```{r consensus-plots, fig.width = 12, fig.height = 9}
(cons$plots$Violin_CD8_TCR) /
  (cons$plots$Primary_vs_Confirm | cons$plots$Bootstrap_ForestPlot)
```

The labels are also written to the object's metadata (prefix `CD8vNK_`), ready
for plotting:

```{r consensus-umap, fig.width = 10, fig.height = 4.5}
grep("^CD8vNK_", colnames(cons$obj@meta.data), value = TRUE)
DimPlot(cons$obj, reduction = "umap.har", group.by = "CD8vNK_consensus") +
  FeaturePlot(cons$obj, reduction = "umap.har", features = "CD8vNK_diff_CD8_TCR") +
  scale_colour_gradient2(low = "#b2182b", mid = "grey90", high = "#2166ac")
```

## Session information

```{r session-info}
sessionInfo()
```
