---
title: "Statistical Analysis: Abundance, Confounders, and Correlations"
author: "Chomchon Uansiri"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Statistical Analysis: Abundance, Confounders, and Correlations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 8,
  fig.height = 5,
  dpi        = 96,
  out.width  = "100%"
)
```

## Overview

Cohort-level questions, such as whether a population expands in disease, must be
tested with the **donor** as the unit of replication, not the cell. Treating
thousands of cells from a few donors as independent observations
(pseudoreplication) produces very small p-values with no biological meaning.
The OHMmy statistics functions aggregate to the donor level before testing:

| Function | Purpose |
|----------|---------|
| `plot_metadata_stats()` | Check that donor-level covariates (age, sex, batch, ...) are balanced across conditions |
| `plot_cell_abundance()` | Test per-donor cell-type proportions across conditions, with automatic test selection and brackets |
| `plot_cluster_distributions()` | Stacked bar charts of counts and proportions for any two metadata columns |
| `plot_gene_pair_correlations()` | Per-cluster correlation between gene pairs |

```{r libraries}
library(Seurat)
library(OHMmy)
library(patchwork)
source("ohmmy-vignette-helpers.R")

pbmc <- load_pbmc3k_ohmmy()   # built in "Data Processing, Integration, and Clustering"
out_dir <- file.path(tempdir(), "OHMmy-statistics")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

### The simulated cohort

The pbmc3k cells are spread across 12 simulated donors in three severity groups.
By construction, monocyte-like cells were over-assigned to Mild and Severe
donors (see `add_mock_cohort()` in `ohmmy-vignette-helpers.R`), so a correct
analysis should find **more CD14 monocytes with increasing severity** and no
other systematic difference.

```{r cohort}
donors <- unique(pbmc@meta.data[, c("donor_id", "severity", "infection_status", "batch", "age", "sex")])
rownames(donors) <- NULL
donors[order(donors$donor_id), ]
```

## 1. Confounder checks with `plot_metadata_stats()`

Before comparing conditions, confirm that the groups do not differ in
covariates that could explain the result. `plot_metadata_stats()` reduces the
metadata to one row per `sample_col`, detects whether each variable is
continuous or categorical, and chooses the test:

* **continuous, 2 groups:** Mann-Whitney U (`"mann_whitney"`) or t-test (`"t_test"`)
* **continuous, 3 or more groups:** Kruskal-Wallis (`"kruskal.test"`) or ANOVA (`"anova"`),
  followed by Dunn's or Tukey post-hoc tests. With `strict_posthoc = TRUE`, the
  post-hoc tests run only when the global test is significant.
* **categorical:** Chi-square (`"chisq"`) or Fisher's exact test (`"fisher"`,
  recommended for small cohorts)

```{r meta-stats, results = "hide"}
meta_res <- plot_metadata_stats(
  seurat_obj         = pbmc,
  sample_col         = "donor_id",
  condition_col      = "severity",
  metadata_vars      = c("age", "sex", "batch"),
  continuous_test_n3 = "kruskal.test",
  categorical_test   = "fisher",
  output_dir         = file.path(out_dir, "metadata"),
  dpi                = 100
)
```

```{r meta-stats-plots, fig.width = 13, fig.height = 4.5}
wrap_plots(meta_res$plots, nrow = 1)
```

Every result is also returned as a table:

```{r meta-stats-tables}
meta_res$stats$age$Global
meta_res$stats$sex$Global
meta_res$stats$batch$Global
```

None of the covariates differ between severity groups, so they are unlikely to
confound the abundance analysis below. The same function handles two-group
designs. Here age is compared between healthy and infected donors with a
t-test:

```{r meta-stats-2grp, results = "hide"}
meta_2grp <- plot_metadata_stats(
  seurat_obj         = pbmc,
  sample_col         = "donor_id",
  condition_col      = "infection_status",
  metadata_vars      = "age",
  continuous_test_n2 = "t_test",
  output_dir         = file.path(out_dir, "metadata"),
  dpi                = 100
)
```

```{r meta-stats-2grp-show, fig.width = 4.5, fig.height = 4.5}
meta_2grp$stats$age$Global
meta_2grp$plots$age
```

## 2. Differential abundance with `plot_cell_abundance()`

`plot_cell_abundance()` computes, for every donor, the fraction of its cells in
each cell type. Missing combinations are filled with zero. It then tests each
cell type across conditions:

* **3 or more conditions:** a global Kruskal-Wallis test (or ANOVA) per cell type,
  then Dunn's (or Tukey's) pairwise tests with `p_adjust` correction. With
  `strict_posthoc = TRUE`, only cell types with a significant global test get
  pairwise brackets.
* **2 conditions:** Mann-Whitney (`"mann_whitney"`), paired Wilcoxon
  (`"wilcoxon_paired"`) or t-test (`"t_test"`).

Brackets are positioned per facet so they never overlap the data.

### Three severity groups

```{r abundance-3, results = "hide"}
p_abund <- plot_cell_abundance(
  seurat_obj       = pbmc,
  sample_col       = "donor_id",
  condition_col    = "severity",
  celltype_col     = "cell_type",
  global_test      = "kruskal.test",
  strict_posthoc   = TRUE,
  p_adjust         = "BH",
  facet_by_cluster = TRUE,
  pairwise_label   = "p.adj",
  y_expand         = 0.35,
  output_dir       = file.path(out_dir, "abundance")
)
```

```{r abundance-3-show, fig.width = 11, fig.height = 10}
p_abund
```

As designed, only **CD14 Mono** passes the global test (Kruskal-Wallis
p ≈ 0.01). Because of `strict_posthoc = TRUE`, it is also the only panel with
pairwise Dunn brackets, and Severe vs Healthy is the significant contrast.
The weaker downward trends in CD4 T, CD8 T and B cells are not an independent
effect. Proportions are compositional, so when monocytes take a larger share of
each Severe donor's cells, every other population's share has to shrink.

### Two conditions, all cell types on one axis

With `facet_by_cluster = FALSE`, all cell types share one panel, with dodged
boxes per condition:

```{r abundance-2, results = "hide"}
p_abund2 <- plot_cell_abundance(
  seurat_obj       = pbmc,
  sample_col       = "donor_id",
  condition_col    = "infection_status",
  celltype_col     = "cell_type",
  pairwise_test    = "mann_whitney",
  facet_by_cluster = FALSE,
  pairwise_label   = "p.signif",
  output_dir       = file.path(out_dir, "abundance")
)
```

```{r abundance-2-show, fig.width = 11, fig.height = 5.5}
p_abund2
```

## 3. Composition overview with `plot_cluster_distributions()`

For a quick descriptive view, `plot_cluster_distributions()` cross-tabulates
any two metadata columns and saves two figures. The first shows proportions in
both directions (cell types within each group, and groups within each cell
type). The second shows absolute counts.

```{r distributions, results = "hide"}
dist_plots <- plot_cluster_distributions(
  seurat_obj  = pbmc,
  cluster_col = "cell_type",
  batch_col   = "severity",
  output_dir  = file.path(out_dir, "distributions"),
  file_prefix = "pbmc3k"
)
```

```{r distributions-show, fig.width = 14, fig.height = 6}
dist_plots$proportions
```

```{r distributions-counts, fig.width = 14, fig.height = 6}
dist_plots$counts
```

## 4. Gene-pair correlations with `plot_gene_pair_correlations()`

Blend plots show co-expression visually. `plot_gene_pair_correlations()` puts a
number on it: for each pair and each cluster, it computes the Pearson,
Spearman or Kendall correlation. Before that, it drops cells that do not
express both genes above the `quantile_thresh` quantile, which reduces the
effect of dropout zeros.

```{r gene-pairs, results = "hide"}
cor_res <- plot_gene_pair_correlations(
  seurat_obj      = pbmc,
  cluster_col     = "cell_type",
  gene_pairs      = list(c("GNLY", "NKG7"), c("CD79A", "MS4A1"), c("S100A8", "S100A9")),
  sample_name     = "pbmc3k",
  cor_method      = "spearman",
  quantile_thresh = 0.01,
  fill_palette    = "kelly",
  output_dir      = file.path(out_dir, "gene_pairs"),
  dpi             = 100,
  verbose         = FALSE
)
```

```{r gene-pairs-show, fig.width = 14, fig.height = 5.5}
cor_res$plot
```

The underlying table also reports how many cells entered each correlation, so
estimates based on only a few cells can be identified:

```{r gene-pairs-table}
head(cor_res$data[order(-cor_res$data$N_Cells), ], 10)
```

## Session information

```{r session-info}
sessionInfo()
```
