---
title: "Differential Expression and Pathway Enrichment"
author: "Chomchon Uansiri"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Differential Expression and Pathway Enrichment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE # Set to FALSE so the vignette builds quickly
)
```

## Introduction

Once you have identified your cell clusters, the next step is discovering what makes them unique biologically. `OHMmy` automates the extraction of cluster markers, pathway enrichment (GSEA and ORA), and visualizes pseudo-bulk differential expression (DESeq2) between experimental conditions.

```{r}

library(Seurat)
library(OHMmy)
library(msigdbr) # Required for pathway databases

# Load your clustered Seurat object
seurat_obj <- readRDS("data/seurat_clustered.rds")
out_dir <- "results/differential_expression/"
```

## Part 1: Finding Top Cluster Markers

The `FindTopMarkersAndHeatmap()` function runs `FindAllMarkers()` to find the defining genes for each cluster, saves the results as a ready-to-use CSV file, and generates a heatmap of the top genes.

```{r}

# 1. Set the active identity to your desired cluster resolution
Idents(seurat_obj) <- "cell_type_annotation"

# 2. Find markers and generate the heatmap
marker_results <- FindTopMarkersAndHeatmap(
  seurat_obj = seurat_obj,
  sample_name = "Cohort_A",
  output_dir_base = paste0(out_dir, "cluster_markers/"),
  plot_format = "jpg",
  onlyPos = FALSE # Set to FALSE to capture both up and down-regulated markers
)

# You can access the saved dataframe directly for the next step
gsea_input_df <- marker_results$markers
```

## Part 2: Pathway Enrichment (GSEA & ORA)

Using the marker dataframe generated in Part 1, we can test if specific biological pathways (like the MSigDB Hallmark gene sets) are enriched in our clusters.

First, we download the Human Hallmark database:

```{r}

m_t2g_hallmark <- msigdbr(species = "Homo sapiens", category = "H")[, c("gs_name", "gene_symbol")]
```

### 2.1 Gene Set Enrichment Analysis (GSEA)

GSEA evaluates the entire ranked list of genes for a cluster (both positive and negative markers) to find coordinated pathway shifts.

```{r}

gsea_results <- run_global_gsea(
  GSEA_df = gsea_input_df, 
  m_t2g = m_t2g_hallmark, 
  title_prefix = "Hallmark",
  top_n_per_direction = 30, # Plot the top 30 up/down pathways per cluster
  output_dir = paste0(out_dir, "GSEA/")
)
```

### 2.2 Over-Representation Analysis (ORA)

ORA evaluates only the *significant* positive markers to see if they overlap significantly with known pathway gene lists.

```{r}

ora_results <- run_global_ora(
  ORA_df = gsea_input_df, 
  m_t2g = m_t2g_hallmark, 
  title_prefix = "Hallmark",
  top_n_per_cluster = 30,
  output_dir = paste0(out_dir, "ORA/")
)
```

## Part 3: Pseudo-bulk Differential Expression (DESeq2)

While `FindAllMarkers()` compares clusters against each other, **DESeq2** compares the *same* cluster across different conditions (e.g., Healthy vs. Disease). `OHMmy` provides wrappers to visualize DESeq2 results natively.

*(Note: Generating the DESeq2 `res` object using pseudo-bulk aggregation is assumed to be completed prior to visualization).*

### 3.1 Volcano Plots

Visualize significant gene expression fold-changes between two conditions.

```{r}

# Define genes you want highlighted on the volcano plot
highlight_genes <- c("IFNG", "GZMB", "TNF", "CX3CR1", "IL7R")

# Generate the plot for Disease vs Control
volcano_plot <- generate_volcano_trio(
  res_obj = res_Disease_vs_Control, # Your DESeq2 results object
  plot_title = "Severe Disease vs Healthy Control", 
  target_genes = highlight_genes
)

# Save the plot
ggsave(paste0(out_dir, "DESeq2/volcano_disease_vs_control.png"), 
       plot = volcano_plot, width = 20, height = 8)
```

### 3.2 Heatmaps of Differentially Expressed Genes

To visualize how the top significant genes fluctuate across individual patients, we use `generate_and_save_heatmap()`.

```{r}

# 1. Transform the count data (Required for heatmaps)
vsd_data <- DESeq2::vst(dds_object, blind = FALSE)

# 2. Setup the annotation column (e.g., Patient Condition)
anno_col <- as.data.frame(colData(dds_object)[, "disease_state", drop = FALSE])
ordered_samples <- rownames(anno_col)[order(anno_col$disease_state)]

# 3. Generate the heatmap
generate_and_save_heatmap(
  res_obj = res_Disease_vs_Control,
  comp_name = "Disease_vs_Control",
  comp_title = "Severe Disease vs Healthy Control",
  vsd_data = vsd_data,
  anno_col = anno_col,
  ordered_samps = ordered_samples,
  n_padj = 50, # Plot top 50 significant genes
  out_dir = paste0(out_dir, "DESeq2/")
)
```
