---
title: "Statistical Analysis: Abundance, Confounders, and Correlations"
author: "Chomchon Uansiri"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical Analysis: Abundance, Confounders, and Correlations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE # Set to FALSE so the vignette builds quickly
)
```

## Introduction

Beyond visualizing clusters and marker genes, rigorous single-cell analysis requires statistical testing. `OHMmy` provides built-in tools to assess clinical cohort balance, calculate differential cell type abundance, and compute statistical correlations between genes.

```{r}

library(Seurat)
library(OHMmy)

# Load your fully annotated Seurat object
seurat_obj <- readRDS("data/seurat_annotated.rds")
out_dir <- "results/statistics/"
```

## Part 1: Confounder Quality Control (Metadata Stats)

Before testing for differences in cell abundance between two conditions (e.g., Healthy vs. Disease), you must ensure your patient cohorts are balanced. If all your diseased patients are significantly older or have different clinical backgrounds than your healthy controls, your results will be confounded.

The `plot_metadata_stats()` function automatically tests categorical and continuous clinical variables across your conditions.

```{r}

plot_metadata_stats(
  seurat_obj = seurat_obj,
  sample_col = "donor_id",       # The column containing individual patient/mouse IDs
  condition_col = "disease_state", # The experimental condition (e.g., Control vs. Infected)
  metadata_vars = c("age", "viral_load", "sex", "batch"), 
  output_dir = paste0(out_dir, "confounder_qc/")
)
```

## Part 2: Cell Type Abundance

To determine if a specific cell population expands or contracts during an infection or treatment, use `plot_cell_abundance()`. This function aggregates cell counts per patient, calculates normalized proportions, and performs statistical testing between your conditions.

```{r}

plot_cell_abundance(
  seurat_obj = seurat_obj,
  sample_col = "donor_id",       
  condition_col = "disease_state", 
  celltype_col = "cell_type_annotation", # The column containing your cluster labels
  
  # Visualization settings
  facet_by_cluster = TRUE,       # Creates individual plots for each cell type
  pairwise_label = "p.adj",      # Display adjusted p-values on the plot
  base_size = 2.5,
  y_expand = 0.35,               # Adds headroom for the statistical significance bars
  
  output_dir = paste0(out_dir, "abundance/")
)
```

## Part 3: Statistical Gene-Pair Correlations

While blend plots can visually show co-expression, sometimes you need hard statistical values to prove that two genes are correlated (or anti-correlated) within a specific cell population.

`plot_gene_pair_correlations()` calculates Pearson or Spearman correlations for specific gene pairs across all your clusters.

```{r}

# Calculate statistical correlations for specific pairs
cor_results <- plot_gene_pair_correlations(
  seurat_obj = seurat_obj,
  cluster_col = "cell_type_annotation",
  
  # List the exact pairs of genes you want to test
  gene_pairs = list(
    c("CD3D", "FCGR3A"),
    c("CD8A", "FCGR3A")
  ),
  
  cor_method = "spearman", 
  sample_name = "Cohort_A",
  output_dir = paste0(out_dir, "correlations/")
)

# You can view the raw correlation table output directly in the console
print(cor_results$data)
```
