---
title: "Diagnostic and Variable Evaluation"
author: "Chomchon Uansiri"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Diagnostic and Variable Evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 8,
  fig.height = 5,
  dpi        = 96,
  out.width  = "100%"
)
```

## Overview

Before trusting a clustering, check what the principal components actually
capture. Two questions matter:

1. **How many PCs carry biological signal?**
2. **Are any PCs driven by technical artefacts** such as mitochondrial or
   ribosomal transcripts, nuclear lncRNAs, sequencing depth or cell cycle?

OHMmy provides five diagnostic functions. Each saves its figures to disk so the
same checks can run unattended on an HPC node.

| Function | Question it answers |
|----------|---------------------|
| `generate_dimheatmaps()` | Which genes and cells drive each PC? |
| `plot_vizdimloadings()` | What are the gene loadings of each PC? |
| `plot_technical_contribution()` | What fraction of each PC's top loadings is technical? |
| `plot_stacked_technical_contribution()` | Does that fraction change with loading depth? |
| `plot_pc_metadata_correlation()` | Do PC scores correlate with QC covariates? |

```{r libraries}
library(Seurat)
library(OHMmy)
source("ohmmy-vignette-helpers.R")

pbmc <- load_pbmc3k_ohmmy()   # built in "Data Processing, Integration, and Clustering"
out_dir <- file.path(tempdir(), "OHMmy-diagnostics")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

The diagnostics use the **pre-integration PCA** (`pca.log`), which still carries
gene loadings.

## 1. Variance and loadings

### Elbow plot

```{r elbow, fig.width = 6, fig.height = 4}
ElbowPlot(pbmc, reduction = "pca.log", ndims = 40)
```

The standard deviation drops steeply over the first 7 PCs and is essentially
flat after PC 10. The 20 PCs used downstream in the processing article are a
conservative choice for pbmc3k.

### Dimension heatmaps

`generate_dimheatmaps()` splits the PCs into windows (`pc_windows`) and saves
one balanced `DimHeatmap` per window. Each window is wrapped in error handling,
so one failed plot does not stop a batch run.

```{r dimheatmaps, results = "hide"}
generate_dimheatmaps(
  seurat_obj  = pbmc,
  sample_name = "pbmc3k",
  reduction   = "pca.log",
  pc_windows  = list(1:9, 10:18),
  output_dir  = out_dir,
  nfeatures   = 20,
  cells       = 300,
  width_in    = 10,
  height_in   = 12,
  res         = 100
)
```

```{r dimheatmaps-show, echo = FALSE}
show_saved(out_dir, "pbmc3k_DimHeatmap_pca.log_PC1-9")
```

PC 1 separates myeloid genes (*LYZ*, *FCN1*, *CST3*) from lymphoid genes
(*IL7R*, *LTB*). Note that *MALAT1* is among its top lymphoid loadings, which
the technical-contribution checks below quantify. PC 2 captures the cytotoxic
NK/CD8 programme against B-cell genes, PCs 3 and 4 platelets, and PC 9 a mixture
of interferon-stimulated and S-phase genes.

### Loading plots

`plot_vizdimloadings()` stitches one `VizDimLoadings` panel per PC into a grid
and returns the names of any windows that failed. An empty vector means
everything was plotted.

```{r vizdim, results = "hide"}
failed <- plot_vizdimloadings(
  seurat_obj  = pbmc,
  sample_name = "pbmc3k",
  reduction   = "pca.log",
  pc_windows  = list(1:6),
  output_dir  = out_dir,
  nfeatures   = 15,
  width_in    = 14,
  height_in   = 9,
  res         = 100,
  ncol        = 3
)
```

```{r vizdim-show}
failed
```

```{r vizdim-img, echo = FALSE}
show_saved(out_dir, "pbmc3k_VizDimLoadings_PC1-6")
```

## 2. Quantifying technical gene contributions

`plot_technical_contribution()` takes the top `n_top_genes / 2` positive and
negative loadings of each PC and computes the **loading-weighted percentage**
that comes from genes matching `technical_keywords`. A dashed line marks the
`cutoff` above which a PC deserves a closer look.

```{r tech-keys}
tech_keys <- c("^MT-", "^RPL", "^RPS", "^IG[HKL]", "MALAT1", "NEAT1", "XIST")
```

```{r tech-contrib, results = "hide"}
plot_technical_contribution(
  seurat_obj         = pbmc,
  sample_name        = "pbmc3k",
  reduction          = "pca.log",
  technical_keywords = tech_keys,
  max_pcs            = 30,
  n_top_genes        = 200,
  cutoff             = 15,
  output_dir         = out_dir
)
```

```{r tech-contrib-show, echo = FALSE}
show_saved(out_dir, "^tech_contrib_pca_log")
```

In pbmc3k every PC stays well below the 15 % cutoff. The largest value is a
few percent on the negative side of PC 1. This is expected, because PCA
loadings only exist for the variable features, and mitochondrial and ribosomal
genes rarely pass HVG selection after QC filtering. In noisier data, such as
stressed tissue, low-quality droplets or nuclei, a PC far above the line is a
strong hint to regress the corresponding covariate.

The percentage can change sharply with the number of genes considered.
`plot_stacked_technical_contribution()` repeats the calculation over a sequence
of depths (`gene_depths`) and facets the result by PC. Bars above the cutoff are
highlighted, so a PC that is "clean" at the top but technical deeper down is
easy to spot.

```{r tech-stacked, results = "hide"}
plot_stacked_technical_contribution(
  seurat_obj         = pbmc,
  sample_name        = "pbmc3k",
  reduction          = "pca.log",
  technical_keywords = tech_keys,
  gene_depths        = seq(0, 300, by = 50),
  max_pcs            = 20,
  cutoff             = 15,
  output_dir         = out_dir,
  plot_width         = 16,
  plot_height        = 10,
  dpi                = 100
)
```

```{r tech-stacked-show, echo = FALSE}
show_saved(out_dir, "^stacked_technical_contrib")
```

## 3. Correlating PCs with QC covariates

`plot_pc_metadata_correlation()` computes Spearman correlations between each
PC's cell embeddings and the metadata columns in `vars_to_test`, then draws a
clustered `pheatmap` with the coefficients printed in each cell. Expression of
single nuclear transcripts, such as *MALAT1*, can be tested by first copying it
into the metadata.

```{r add-malat1}
pbmc$MALAT1_expr <- FetchData(pbmc, vars = "MALAT1", layer = "data")[, 1]
```

```{r pc-cor, results = "hide"}
plot_pc_metadata_correlation(
  seurat_obj     = pbmc,
  sample_name    = "pbmc3k",
  vars_to_test   = c("nCount_RNA", "nFeature_RNA", "pct_counts_mt", "percent.ribo",
                     "S.Score", "G2M.Score", "MALAT1_expr"),
  reduction      = "pca.log",
  n_pcs          = 20,
  output_dir     = out_dir,
  plot_width_in  = 13,
  plot_height_in = 5,
  res_dpi        = 100
)
```

```{r pc-cor-show, echo = FALSE}
show_saved(out_dir, "^corr_pc_pbmc3k")
```

## Interpreting the diagnostics

* A PC whose top loadings are dominated by `MT-`, `RPL`/`RPS` or lncRNA genes,
  and whose scores correlate strongly (|rho| > 0.5) with `pct_counts_mt`,
  `percent.ribo` or `nCount_RNA`, is technical. Regress that covariate in
  `ProcessSeuratLOG(vars_to_regress = ...)`, switch to `ProcessSeuratSCT()`, or
  leave the PC out of `dims`.
* Ribosomal genes often load on lymphocyte-versus-myeloid PCs in PBMCs simply
  because naive T cells are ribosome-rich. Before removing such a PC, check
  whether it also separates known cell types, for example with the dimension
  heatmaps above.
* Correlation with `S.Score` or `G2M.Score` indicates a cycling compartment.
  Regress the cell-cycle difference only when proliferation is not of interest.

## Session information

```{r session-info}
sessionInfo()
```
