---
title: "Comprehensive single-cell workflow with OHMmy"
author: "Chomchon Uansiri"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Comprehensive single-cell workflow with OHMmy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 8,
  fig.height = 5,
  dpi        = 96,
  out.width  = "100%"
)
```

## Introduction

**OHMmy** provides reproducible wrappers for a complete Seurat v5 scRNA-seq
analysis: ambient-RNA removal with SoupX, normalisation and batch integration,
diagnostics of technical noise, publication-ready visualisation, donor-aware
statistics, and pathway enrichment. Every plotting function saves a
timestamped, high-resolution file **and** returns the underlying ggplot or
data, so OHMmy fits both batch scripts on an HPC cluster and interactive
analysis.

This article gives a short, runnable tour of the main steps on one dataset. The
other articles cover each module in depth:

| Article | Covers |
|---------|--------|
| [Ambient RNA Decontamination with SoupX](OHMmy-soupx-workflow.html) | Reading Cell Ranger output, estimating and removing ambient RNA |
| [Data Processing, Integration, and Clustering](OHMmy-processing-clustering.html) | LogNormalize / SCTransform pipelines, integration, resolution sweeps |
| [Diagnostic and Variable Evaluation](OHMmy-diagnostics-evaluation.html) | PC heatmaps and loadings, technical-gene contribution, PC-covariate correlation |
| [Advanced Visualization and Marker Analysis](OHMmy-advanced-visualization.html) | UMAP overlays, density plots, clustered dot plots, blend plots, CD8/NK resolution |
| [Statistical Analysis](OHMmy-statistical-analysis.html) | Confounder checks, differential abundance, compositions, gene-pair correlations |
| [Differential Expression and Pathway Enrichment](OHMmy-differential-expression.html) | Markers, GSEA, ORA, pseudo-bulk DESeq2 visualisation |

## Function index

```{r function-index, echo = FALSE}
idx <- data.frame(
  Module = c(
    rep("Ambient RNA (SoupX)", 10),
    rep("Processing", 4),
    rep("Diagnostics", 5),
    rep("Visualization", 12),
    rep("Lineage resolution", 1),
    rep("Statistics", 4),
    rep("Markers & DE", 6)
  ),
  Function = c(
    "get_sample_names", "load_counts", "create_soup_channels", "create_seurat_for_clustering",
    "estimate_contamination", "create_final_seurat", "prepare_soupx_inputs",
    "run_soupx_post_clustering", "process_soupx_samples", "addSoupXMetaToSeurat",
    "ProcessSeuratLOG", "ProcessSeuratSCT", "ClusterAndUMAP", "CleanSeuratReductions",
    "generate_dimheatmaps", "plot_vizdimloadings", "plot_technical_contribution",
    "plot_stacked_technical_contribution", "plot_pc_metadata_correlation",
    "make_cluster_palette", "PlotDimByFactors", "plot_combined", "plot_violin_qc_single",
    "extract_binned_expression", "plot_dot_dendro", "plot_dot_dendro_split",
    "plot_dot_dendro_multi", "plot_blend_nebulosa", ".blend_feature_plot_v5",
    "plot_gene_markers_with_dendro", "plot_split_dotplots_by_gene_cluster",
    "score_consensus",
    "plot_metadata_stats", "plot_cell_abundance", "plot_cluster_distributions",
    "plot_gene_pair_correlations",
    "FindTopMarkersAndHeatmap", "run_global_gsea", "run_global_ora",
    "generate_volcano_trio", "get_top_mixed_genes", "generate_and_save_heatmap"
  ),
  Article = c(
    rep("SoupX", 10), rep("Processing", 4), rep("Diagnostics", 5),
    rep("Advanced visualization", 10), rep("Differential expression", 2),
    "Advanced visualization", rep("Statistics", 4), rep("Differential expression", 6)
  )
)
idx$Function <- paste0("`", idx$Function, "()`")
knitr::kable(idx, row.names = FALSE)
```

## The example dataset

All articles use the 10x Genomics **3k PBMC** dataset (pbmc3k). It is
downloaded once and cached. The helper file `ohmmy-vignette-helpers.R` also
spreads the cells across a **simulated cohort of 12 donors** (Healthy, Mild and
Severe, in two sequencing batches), so that the cohort-level functions can be
demonstrated. Monocyte-like cells were deliberately over-assigned to Mild and
Severe donors. *The donor, severity, batch, age and sex columns are synthetic.*

```{r libraries}
library(Seurat)
library(OHMmy)
library(patchwork)
source("ohmmy-vignette-helpers.R")

out_dir <- file.path(tempdir(), "OHMmy-comprehensive")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

## 1. Load, QC and annotate the cohort

Ambient-RNA removal with SoupX would normally come first. It needs the raw
droplet matrices, so it is demonstrated separately in the SoupX article.

```{r load}
pbmc <- CreateSeuratObject(pbmc3k_counts(), project = "pbmc3k", min.cells = 3, min.features = 200)
pbmc[["pct_counts_mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
pbmc[["percent.ribo"]]  <- PercentageFeatureSet(pbmc, pattern = "^RP[SL]")
pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & pct_counts_mt < 5)
pbmc <- add_mock_cohort(pbmc)
table(pbmc$severity)
```

## 2. Normalise, integrate and cluster

```{r process, results = "hide"}
pbmc <- ProcessSeuratLOG(
  pbmc,
  batch_col             = "batch",
  vars_to_regress       = "pct_counts_mt",
  tcr_bcr_patterns      = "^TR[ABDG]V|^IG[HKL]V",
  reduction_name        = "pca.log",
  integration_method    = "HarmonyIntegration",
  integration_reduction = "harmony.log",
  verbose               = FALSE
)

clus <- ClusterAndUMAP(
  pbmc,
  sample_name         = "pbmc3k",
  reduction           = "harmony.log",
  dims                = 1:20,
  umap_name           = "umap.har",
  graph_name          = "har_snn",
  cluster_prefix      = "har_snn_res.",
  cluster_resolutions = seq(0.2, 1.2, by = 0.2),
  final_resolution    = 0.6,
  plot_dir            = out_dir,
  dpi                 = 100
)
pbmc <- clus$seurat
pbmc$seurat_clusters <- pbmc$har_snn_res.0.6
pbmc <- annotate_pbmc(pbmc, cluster_col = "seurat_clusters")
```

```{r umap, fig.width = 11, fig.height = 4.5}
clus$clustree + DimPlot(pbmc, reduction = "umap.har", group.by = "cell_type",
                        label = TRUE, repel = TRUE) + NoLegend()
```

## 3. Check the principal components for technical noise

```{r diagnostics, results = "hide"}
plot_technical_contribution(
  pbmc, sample_name = "pbmc3k", reduction = "pca.log",
  max_pcs = 20, n_top_genes = 200, output_dir = out_dir
)
```

```{r diagnostics-show, echo = FALSE}
show_saved(out_dir, "^tech_contrib_pca_log")
```

## 4. Visualise markers

```{r dotplot, results = "hide"}
feature_df <- data.frame(
  group   = paste0("PBMC_", rep(names(pbmc_marker_sets), lengths(pbmc_marker_sets))),
  feature = unlist(pbmc_marker_sets, use.names = FALSE)
)
p_dot <- plot_dot_dendro(pbmc, meta_col = "cell_type", feature_df = feature_df,
                         prefix = "PBMC", pct_threshold = 10, output_dir = out_dir)
```

```{r dotplot-show, fig.width = 12, fig.height = 7.5}
p_dot
```

```{r blend, results = "hide"}
blend <- plot_blend_nebulosa(pbmc, cluster_col = "cell_type",
                             gene_pairs = list(c("CD14", "FCGR3A")),
                             reduction = "umap.har", width = 24, dpi = 72,
                             output_dir = out_dir, verbose = FALSE)
```

```{r blend-show, fig.width = 24, fig.height = 4.2}
blend$plot
```

## 5. Test for differences in cell-type abundance

```{r abundance, results = "hide"}
p_abund <- plot_cell_abundance(
  pbmc,
  sample_col     = "donor_id",
  condition_col  = "severity",
  celltype_col   = "cell_type",
  pairwise_label = "p.adj",
  y_expand       = 0.35,
  output_dir     = out_dir
)
```

```{r abundance-show, fig.width = 11, fig.height = 10}
p_abund
```

The test recovers the simulated effect: CD14 monocytes increase with severity.

## 6. Markers and pathways

```{r markers, results = "hide"}
Idents(pbmc) <- "cell_type"
mk <- FindTopMarkersAndHeatmap(pbmc, sample_name = "pbmc3k", top_n = 5, onlyPos = FALSE,
                               output_dir_base = out_dir, height = 10, dpi = 72)

hallmark <- msigdbr::msigdbr(species = "Homo sapiens", collection = "H")[, c("gs_name", "gene_symbol")]
gsea <- run_global_gsea(
  GSEA_df      = dplyr::mutate(mk$markers, cluster = as.character(cluster)),
  m_t2g        = hallmark,
  output_dir   = file.path(out_dir, "gsea"),
  title_prefix = "Hallmark"
)
```

```{r gsea-show, echo = FALSE}
show_saved(file.path(out_dir, "gsea"), "Hallmark_GSEA_Clustered_Dendro_Dotplot")
```

## Where to go next

* Clean your raw Cell Ranger output first: see the SoupX article.
* Validate your PC choice before clustering: see the Diagnostics article.
* For donor-aware differential expression, pseudo-bulk DESeq2 and its
  visualisations are covered in the Differential Expression article.

## Session information

```{r session-info}
sessionInfo()
```
